\chapter{Results}
\label{chap:results}

This chapter presents the experimental results obtained during the course of this study. A total of 285 experiments were documented across 290 experiment IDs, conducted between November 2021 and June 2025, spanning CARLA simulator setup, image classification, and autonomous driving model development.

Initial experiments (1-8) established the CARLA simulation environment with traffic generation, dynamic weather conditions, and vehicle control systems. Image classification experiments (11-28) systematically evaluated CNN accuracy degradation against 12 perturbation types on MNIST datasets, including development of a perturbed MNIST dataset for controlled noise analysis.

The research progressed through three phases. The early phase (experiments 1-50) focused on CNN baseline performance and simulation infrastructure development. The middle phase (experiments 51-150) introduced Vision Transformer architectures for image classification and autonomous driving, examining parameter scaling and pre-trained model adaptation. The advanced phase (experiments 151-290) incorporated Vision Language Models, applying multimodal architectures to the problem of steering prediction under noisy conditions.

Key experimental themes included noise impact analysis (47 experiments), infrastructure and hardware evaluation (13 experiments), and systematic comparison of CNN, ViT, and VLM architectures across multiple datasets. The methodology development progressed from image classification to regression-based steering prediction to classification-based approaches using discrete steering angle values discretized into bins. Autonomous driving experiments constituted approximately 70\% of the total research effort (201/285 experiments), with the CARLA simulator serving as the primary experimental platform. The experimental sequence shows progression from traditional computer vision approaches through convolutional models and transformer models to vision and multimodal large language models. The experiment logs are presented in a separate appendix.

The following sections present results organized according to the thesis objectives:

\begin{itemize}

\item Distance-based Distribution Shift Detection
\item Cross-architectural Validation on Standard Datasets
\item Accuracy Degradation Characterisation
\item Synthetic Dataset Creation for Autonomous Driving
\item Autonomous Driving Safety Application
\end{itemize}
